---
title: "Heart Disease Prediction"
author: "Nazman, Ezgi"
date: "2024-10-17"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Cleveland Heart Disease Data 

## Dataset Overview:

1.  **Source:** UCI Machine Learning Repository - Heart Disease Dataset

2.  **Purpose:** The primary goal is to predict the presence or absence of heart disease in a patient, based on a set of medical attributes.

3.  **Original Data:** The dataset was originally collected by the Cleveland Clinic Foundation.

## Key Characteristics of Cleveland Data:

1.  **Total Instances:** 303 rows

2.  **Total Attributes:** 14 columns (13 features and 1 target variable)

3.  **Missing Values:** Some instances contain missing data, represented by ?.

## Attribute Information:

The dataset contains 13 medical attributes (or features) and 1 target variable that indicates the presence of heart disease.

1.  **age:** Age of the patient in years.

2.  **sex:** Gender of the patient (1 = Male, 0 = Female).

3.  **cp (chest pain type):** **1=** Typical angina.
    **2=** Atypical angina.
    **3=** Non-anginal pain.
    **4=** Asymptomatic.

4.  ***trestbps:*** Resting blood pressure (in mm Hg) on admission to the hospital.

5.  **chol:** Serum cholesterol level (in mg/dL)**.**

6.  **fbs** (fasting blood sugar):Whether the fasting blood sugar is greater than 120 mg/dL (1 = True, 0 = False).

7.  **restecg** (resting electrocardiographic results): **0=** Normal.
    **1=** Having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of \> 0.05 mV).
    **2=**Showing probable or definite left ventricular hypertrophy by Estes' criteria.

8.  **thalach:** Maximum heart rate achieved during exercise.

9.  **exang:** Exercise-induced angina (**1** = Yes, **0** = No).

10. **oldpeak:** ST depression induced by exercise relative to rest (numeric value measured in mm).

11. **slope:**The slope of the peak exercise ST segment: **1=** Upsloping.
    **2=** Flat.
    **3=** Downsloping.

12. **ca:** Number of major vessels (0-3) colored by fluoroscopy (higher values indicate more blocked vessels).

13. **thal:** A blood disorder called thalassemia: **3 =** Normal.
    **6** = Fixed defect.
    **7** = Reversible defect.

14. **num** (target variable): Diagnosis of heart disease (angiographic disease status).
    Originally a categorical variable ranging from 0 to 4.
    **0**= No heart disease.
    **1, 2, 3, 4**= Different levels of heart disease severity.
    In this study, this is simplified into a binary classification (**0** = No heart disease, **1** = Presence of heart disease).

```{r}
# Define the URL for the dataset
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"

# Define column names
col_names <- c('age', 'sex', 'cp', 'trestbps', 'chol', 'fbs', 'restecg', 'thalach', 
               'exang', 'oldpeak', 'slope', 'ca', 'thal', 'num')

# Load the dataset directly from the URL
heart_data <- read.table(url, sep = ",", col.names = col_names, na.strings = "?")

# View the first six rows of the dataset

head(heart_data)
```

```{r}
# Load necessary libraries for plotting
library(ggplot2)
library(gridExtra)
library(reshape2)
```

```{r}
# Select the relevant columns for correlation analysis
cor_data <- heart_data[, c('age', 'trestbps', 'chol', 'thalach', 'oldpeak')]

# Calculate the correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs")

# Melt the correlation matrix for ggplot
cor_melted <- melt(cor_matrix)

# Create the heatmap
cor_heatmap <- ggplot(data = cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display the heatmap
print(cor_heatmap)
```

```{r}
#Histogram plot showing the distribution of patient ages
age_plot <- ggplot(heart_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  theme_minimal()
```

```{r}
#Bar plot showing the count of male and female patients
sex_plot <- ggplot(heart_data, aes(x = factor(sex))) +
  geom_bar(fill = "coral", color = "black") +
  labs(title = "Distribution of Sex", x = "Sex (0 = Female, 1 = Male)", y = "Count") +
  theme_minimal()
```

```{r}
#Bar plot showing different types of chest pain reported by the patients.
cp_plot <- ggplot(heart_data, aes(x = factor(cp))) +
  geom_bar(fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Chest Pain Type", x = "Chest Pain Type", y = "Count") +
  theme_minimal()
```

```{r}
#Box plot comparing the resting blood pressure of patients with and without heart disease.
bp_plot <- ggplot(heart_data, aes(x = factor(num), y = trestbps)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Resting Blood Pressure by Heart Disease", x = "Heart Disease (0 = No, 1 = Yes)", y = "Resting Blood Pressure") +
  theme_minimal()
```

```{r}
#Box plot comparing the resting blood pressure of patients with and without heart disease.
chol_plot <- ggplot(heart_data, aes(x = factor(num), y = chol)) +
  geom_boxplot(fill = "lightpink", color = "black") +
  labs(title = "Cholesterol Level by Heart Disease", x = "Heart Disease (0 = No, 1 = Yes)", y = "Cholesterol") +
  theme_minimal()  
```

```{r}
# Arrange the above plots in a grid for visualization
grid.arrange(age_plot, sex_plot, cp_plot, bp_plot, chol_plot, ncol = 2)
```

### Data Preparation for binary Logistic Regression

Here the num column indicates the presence of heart disease (0 = No, 1 = Yes).

```{r}
# Convert 'num' column to binary: 0 for no heart disease, 1 for any level of heart disease
heart_data$num <- ifelse(heart_data$num > 0, 1, 0)

# Remove rows with missing values
heart_data <- na.omit(heart_data)

# Check the structure of the dataset
str(heart_data)
```

### Model Fitting

**1.** Significant predictors (marked with \* or \*)**,** such as sex, chest pain type, blood pressure, maximum heart rate, exercise-induced angina, and thalassemia.

**2.** Coefficients indicate how each variable influences the likelihood of heart disease.

**3**.
The p-values show the statistical significance of each predictor in the model.

```{r}
# Fit a binary logistic regression model to predict heart disease
model <- glm(num ~ age + sex + cp + trestbps + chol + fbs + restecg + thalach + 
             exang + oldpeak + slope + ca + thal, 
             data = heart_data, family = binomial)

# View the summary of the model
summary(model)
```

#### Key interpretation of the results:

1.  The intercept is significant, suggesting a meaningful baseline log-odds for heart disease when all predictors are zero.

2.  Sex (male) is a significant predictor, with males being more likely to have heart disease (p-value = 0.00723).

3.  Chest pain type (cp) is highly significant (p-value = 0.00259), with more severe types of chest pain strongly associated with heart disease.

4.  Resting blood pressure (trestbps) is also significant, showing that higher blood pressure slightly increases the risk (p-value = 0.02504).

5.  Maximum heart rate achieved (thalach) is significant (p-value = 0.04327), with lower maximum heart rates increasing heart disease risk.

6.  Exercise-induced angina (exang) is significant (p-value = 0.02506), increasing the likelihood of heart disease.

7.  The number of major vessels colored by fluoroscopy (ca) is one of the most significant predictors (p-value = 1.78e-06), showing a strong relationship with heart disease.

8.  Thalassemia (thal) is also highly significant (p-value = 0.00061), with certain types of thalassemia being associated with higher risk.

#### Non-significant predictors:

Age, cholesterol, fasting blood sugar, resting electrocardiographic results, ST depression (oldpeak), and slope of the ST segment were not significant.
Note that this might be due to characteristics specific to this dataset.

#### Model fit:

1.  The model provides a good fit, with a substantial reduction in deviance (from Null deviance of 409.95 to Residual deviance of 204.69).

2.  AIC (Akaike Information Criterion) = 232.69, suggesting that this model strikes a balance between goodness of fit and complexity.
